Background: Multifunctional Lipid Probes

### 2025-05-14 Background/MultifunctionalLipidProbesOverview.qmd
updating some of Carsten's edits and some other small changes

Functionalized lipid probes are at the intersection of several niche biotechnologies. This overview will serve as a summary of the capabilities, nuances, and limitations of their utility. Additional resources and published review articles will be referenced and linked below for greater detail.

@@

At its core, biology is a complex network of chemical interactions. Every biological process—from metabolism to signaling—depends on the physical interactions between molecules, including proteins, nucleic acids, and lipids. While proteins have long been the primary focus of interaction studies, lipids play an equally crucial yet often under-appreciated role in cellular function. Therefore, there is a need for detailed knowledge of lipid-protein interactions (the lipid interactome), in healthy and diseased cells and tissue.

@@

Lipids are primarily organized within membrane bilayers, where their interactions define critical membrane properties such as rigidity, fluidity, and resistance to detergents. By studying lipid-lipid interactions and lipid interactomes, researchers will gain deeper insight into how these molecular interactions influence local and global membrane architecture and dynamics. Understanding these interactions and interactomes is essential for unraveling fundamental processes like vesicle trafficking, cell signaling, and membrane compartmentalization.

@@

Beyond their structural roles, many lipids have been increasingly recognized as key signaling molecules. For example, the sphingolipid sphingosine-1-phosphate (S1P) plays a pivotal role in immune function. S1P binds to a specific family of receptors, triggering T cell migration (extravasation) and guiding their movement toward sites of infection. Rapidly turned-over phosphatidylinositol 3,4,5-*tris*phosphate triggers cell survival, growth, movement, and proliferation. These examples  highlights how lipid signaling can affect cellular behavior and physiological responses.

@@

The study of lipid biology has been greatly advanced by synthetic lipid analogs. However, it may not be immediately clear why such analogs are essential research tools. Unlike proteins and nucleic acids, lipids present unique challenges that make traditional genetic tools less effective in studying their function. In addition, synthetic lipid tools are designed to provide a defined composition of fatty acids and carbon backbones connected to a specific head group while genetic and pharmacological methods are more likely to produce a diverse group of lipid species.

@@

To overcome these challenges, researchers have developed synthetic analogs of native lipids. These modified lipids allow for precise investigation of lipid function while preserving their essential biological properties. The key modifications used in this approach include a photo-crosslinking group (usually a diazirine), an alkyne group for click chemistry, and a photo-removable (cage) group for preventing premature probe metabolism.

#### Diazirine group {#sec-Diazirine}

@@

##### Key properties of diazirine crosslinking:

@@

Among the various crosslinking motifs, the diazirine ring is the smallest, making it the least likely to interfere with a lipid’s natural interactions. Upon exposure to 355 nm UV light, the diazirine ring forms a reactive carbene radical, which rapidly covalently bonds with nearby molecules. This reaction is highly versatile, as the carbene can insert into carbon-hydrogen, sulfur-hydrogen, nitrogen-hydrogen, oxygen-hydrogen, or double bonds, allowing for broad coverage of potential interaction partners[@haberkantFatFabulousBifunctional2014].Despite these numerous options, reaction with glutamate seems to be modestly preferred [@westDesignEvaluationCyclobutane2022].

The small size and high reactivity of diazirine crosslinking make it an invaluable tool for capturing lipid-protein interactions while minimizing perturbations to native lipid function.

##### The effect of diazirine placement on crosslinking targets {#sec-DiazirinePlacement}

@@

@@

##### A versatile bioorthogonal handle

@@

The terminal alkyne is particularly useful because it undergoes azide-alkyne cycloaddition, the prototypical click chemistry reaction. This reaction is highly specific, rapid, and bio-orthogonal, meaning it does not interfere with natural cellular processes. By reacting the alkyne with an azide-containing probe, researchers can introduce fluorescent markers for imaging or affinity tags for isolating lipid-protein complexes. The reaction requires catalysis by copper (I) ions and is therefore usually not suitable for intact cells and tissues but works well in fixed or lysed samples.

@@

Another important use is affinity-based enrichment. Lipids containing a terminal alkyne can be captured using azide-functionalized beads, enabling the isolation of lipid-protein complexes after photo-crosslinking for analysis by mass spectrometry. This method has been instrumental in identifying proteins that interact with specific lipid species, shedding light on lipid signaling pathways and metabolic networks.

#### Coumarin Photocage {#sec-Coumarin}
Lipids play an essential role in cellular signaling which often relies on transient abundance of lipid species. Hence, studying lipids in intact cells presents unique challenges due to their highly dynamic metabolism. Unlike proteins, which are relatively stable once synthesized, many bioactive lipids are continuously synthesized, modified, interconverted, and degraded as part of an extensive and tightly regulated lipid metabolic network. Even lipids so far not associated with cellular signaling show rapid turnover, which makes it difficult to study the effects of a specific lipid species in isolation, as the lipid of interest may be enzymatically altered before it can exert its intended biological function.

One strategy to overcome this challenge is using photocaged lipid analogs, which remain biologically inert until exposed to a specific wavelength of light. Irradiation with this wavelength "uncages" the lipid derivative within seconds to reveal its native structure. This allows researchers to control the timing and location of lipid activation, enabling targeted investigation of lipid function with high spatiotemporal resolution.

##### A light-activated protection group

One of the most effective tools for photocaging lipids is based on 7-diethylamino-4-methylene coumarin, a photolabile protection group that can be removed upon exposure to 400 nm light. When attached to a lipid’s headgroup, the coumarin moiety temporarily blocks interactions with proteins, preventing the lipid from participating in its normal biological pathways. However, upon photoactivation, the coumarin group is cleaved, restoring the lipid to its native structure and allowing it to engage with its biological targets.

This approach is useful for studying lipid-protein interactions, as uncaging the lipid in a controlled manner ensures that protein binders recognize the native lipid structure only when and where the researcher intends.

An alternative to coumarin-based caging groups are o-nitrobenzyl.

@@

The power of coumarin-caged lipids has been demonstrated in studies examining intracellular lipid signaling. For example, in 2015, Höglinger et al. used a caged sphingosine analog to investigate the role of sphingosine in calcium signaling. They found that upon photoactivation of the caged sphingosine, lysosomal calcium release was triggered, revealing a crucial link between sphingosine metabolism and intracellular calcium dynamics [@hoglingerIntracellularSphingosineReleases2015a]. This discovery significantly expanded the understanding of the sphingosine signaling cascade and its downstream effects.

@@

##### An Additional Benefit: Coumarin’s Inherent Fluorescence

A unique advantage of using coumarin as a photocaging group is its intrinsic fluorescence. This means that before the lipid is even uncaged, its cellular incorporation can be directly visualized using a fluorescence microscope. By tracking fluorescence intensity and distribution, researchers can quickly determine whether a functionalized lipid has successfully intercalated into the target cell’s membrane or accumulated in specific subcellular compartments.

This dual functionality—serving as both a caging group and a fluorescent probe—makes coumarin a useful tool for studying lipid localization, uptake, and bioactivity.

These modifications stabilize protein-lipid interactions and enable the selective enrichment of lipid-bound proteins, all while maintaining tight spatial and temporal control. Each of these tools is described in further detail below.

## Limitations of multifunctionalized lipid probes

* While significant effort has minimized the effect that each modification has on the biophysical properties of the lipid analogs, it remains the case that adding a diazirine and a terminal alkyne will necessarily affect their physical properties -- perhaps affecting sub-cellular localization and binding. 
* The coumarin cage has been repeatedly shown to protect lipid analogs from premature metabolism, but it should be recognized that this comes with some distinct changes to the behavior of the native lipid. 
  - For example, when a cell is treated with a photocaged lipid probe, the probe intercalates into every membrane in the cell -- including membranes which the native lipid would not enter. This may lead to the identification of non-biologically-relevant hits. 
* Similarly, the diazirine and alkyne groups may have minor effects on the biophysical properties of the lipid probe as it interacts with other lipids and proteins within the membrane bilayer. This may be more relevant for proteins which preferentially bind to the acyl chain. 
  - For example, Some work has been done to demonstrate that the placement of the diazirine ring along the length of the acyl chain can have impacts on the interactome of the lipid probe[@farleyTrifunctionalFattyAcid2024]. Farley et al (2024) demonstrated this using trifunctional Fatty Acid probes by placing the diazirine either 3 carbons from the terminus (8-3 FA) or 10 carbons from the terminus (1-10 FA). They identified several proteins with distinct enrichment profiles, including VDAC1 and VDAC2 (which are preferentially enriched to the 8-3 FA probe), and ACAD9 and UBB (which are preferentially enriched to the 1-10 FA probe). 
* Quantification of +UV versus -UV
  - The details of the quantitative mass spectrometry are describe in our [Overview of Proteomics Methods](ProteomicsUsingMultifunctionalProbes.qmd), however, we hope to succinctly describe that quantitative mass spectrometry techniques necessarily require normalization at sample input level -- thereby exaggerating the abundance of proteins in the -UV control samples. 
  - It should be noted, therefore, that there is little meaning to a protein which is "enriched" toward the -UV condition. For this reason, all proteins which are "depleted" following irradiation with UV have been designated as "no hit" in our analyses.

>>>>

Functionalized lipid probes are a toolset intended to fuse several niche biotechnologies. This overview will serve as a summary of the capabilities, nuances, and limitations of their utility. Please find additional resources and published primary and review articles below for greater detail.

@@

At its core, biology is a complex network of chemical interactions. Every biological process—from metabolism to signaling—depends on the physical interactions between molecules, including proteins, nucleic acids, and lipids. While proteins have long been the primary focus of interaction studies, lipids play an equally crucial yet often under-appreciated role in cellular function. Therefore, there is a need for detailed knowledge of lipid-protein interactions (the lipid interactome). To learn about diseases and to find new therapeutic targets, in the comparison of the lipid interactomes of healthy and diseased cells and tissue is currently developing into a new field of lipid biology.

@@

Lipids are primarily organized within membrane bilayers, where their interactions define critical membrane properties such as rigidity, fluidity, and resistance to detergents. By studying lipid-lipid interactions and lipid interactomes, researchers will gain deeper insight into how these molecular interactions influence local and global membrane architecture and dynamics. Understanding these interactions and interactomes is essential for unraveling fundamental processes such as vesicle trafficking, cell division, cell signaling, and membrane compartmentalization.

@@

Beyond their structural roles, many lipids have been increasingly recognized as key signaling molecules. For example, the sphingolipid sphingosine-1-phosphate (S1P) plays a pivotal role in immune function. S1P binds to a specific family of receptors, triggering T cell migration (extravasation) and guiding their movement toward sites of infection. Rapidly turned-over phosphatidylinositol 3,4,5-*tris*phosphate triggers cell survival, growth, movement, and proliferation. These examples  highlight how lipid signaling can affect cellular behavior and physiological responses.

### Lipids in Disease

Many diseases are caused by or affiliated with a lack or an accumulation of lipids, e.g. in atherosclerosis of in lipid storage diseases. However, recent evidence shows that in many if not most disease states specific lipid species are up- or down-regulated. A good example are viral and bacterial infections. If we learn about the effects of these altered lipids, we might identify new biochemical mechanisms essential for disease development and hence new therapeutic targets.

@@

The study of lipid biology has been greatly advanced by synthetic lipid analogs. However, it may not be immediately clear why such analogs are essential research tools. Unlike proteins and nucleic acids, lipids present unique challenges that make traditional genetic tools less effective in studying their function. In addition, synthetic lipid tools are designed to provide a defined composition of fatty acids and carbon backbones connected to a specific head group while genetic and pharmacological methods are more likely to produce a diverse group of lipid species. The lipid binding pocket of a protein might only in some cases be a great therapeutic target while the function of the lipid binding protein might well be.

@@

To overcome the challenges outlined above, researchers have developed synthetic analogs of native lipids. These modified lipids allow for precise investigation of lipid function while preserving their essential biological properties. The key modifications used in this approach include one to three functional modifications, namely a photo-crosslinking group (usually a diazirine), an alkyne group for click chemistry, and a photo-removable (cage) group for preventing premature probe metabolism.

#### Diazirine Group {#sec-Diazirine}

@@

##### Key Properties of Diazirine Crosslinking:

@@

Among the various crosslinking motifs, the diazirine ring is the smallest, making it the least likely to interfere with a lipid’s natural interactions. Still, the introduction of diazirine into a fatty acid side chain will slightly decrease hydrophobicity and lipid ordering. Upon exposure to 355 nm UV light, the diazirine ring forms a reactive intermediate, which rapidly form covalent bonds with nearby molecules. This reaction is highly versatile, as reactive intermediates such as carbenes can insert into nearby carbon-hydrogen, sulfur-hydrogen, nitrogen-hydrogen, oxygen-hydrogen, or double bonds, allowing for broad coverage of potential interaction partners1. Despite these numerous options, reaction with glutamate seems to be modestly preferred2.
The small size and high reactivity of diazirine crosslinking make it an invaluable tool for capturing lipid-protein interactions while minimizing perturbations to native lipid function.

##### The Effect of Diazirine Placement on Crosslinking Targets {#sec-DiazirinePlacement}

@@

Increasingly, lipid derivatives with the diazirine located on or near the lipid head group become available. Hence, many additions to the lipid interactome database now show interactions with proteins located at the membrane interface or cytosolic proteins. Future comparison of intramembrane versus membrane interface crosslinking will provide a more global view on lipid interactomes.

@@

##### A Versatile Bioorthogonal Handle

@@

The terminal alkyne is particularly useful because it undergoes copper-catalyzed azide-alkyne cycloaddition, the prototypical click chemistry reaction. This reaction is highly specific, rapid, and bio-orthogonal, meaning it does not produce reaction products with natural cellular molecules. . By reacting the alkyne with an azide-containing probe, researchers can introduce fluorescent markers for imaging or affinity tags for isolating lipid-protein complexes. The reaction requires catalysis by copper (I) ions and is therefore usually not suitable for intact cells and tissues but works well in fixed or lysed samples.

@@

Another important use is affinity-based enrichment. Lipids containing a terminal alkyne can be captured using azide-functionalized beads (or resins), enabling the isolation of lipid-protein complexes after photo-crosslinking for analysis by mass spectrometry. This method has been instrumental in identifying proteins that interact with specific lipid species, shedding light on lipid signaling pathways and metabolic networks. An alternative is to covalently attach biotin to the lipid-protein conjugates and the use of streptavidin beads to wash and isolate them.

#### The Photo-activatable Group (Cage) {#sec-Coumarin}

Essential roles of lipids in cellular signaling often rely on a transient abundance of lipid species, in some cases on the sub-minute scale. Hence, studying lipids in intact cells presents unique challenges due to their highly dynamic metabolism. Unlike proteins, which are relatively stable once synthesized, many bioactive lipids are continuously synthesized, modified, interconverted, and degraded as part of an extensive and tightly regulated lipid metabolic network. Even lipids so far not associated with cellular signaling show rapid turnover, which makes it difficult to study the effects of a specific lipid species in isolation, as a lipid of interest artificially introduced to cells may be enzymatically altered before it can exert its intended biological function.

One strategy to overcome this challenge is using photo-activatable (caged) lipid derivatives. These  remain biologically inert until exposed to a specific wavelength of light. Irradiation releases the biologically active lipid derivative within seconds. Subcellular illumination allows researchers to control the timing and location of lipid levels, enabling targeted investigation of lipid function with high spatiotemporal resolution.

##### A Light-Activated Protection Group

There is a limited but significant number of photo-activatable protecting groups that have been used successfully in biological experiments, One of the most effective tools for photocaging lipids is based on 7-diethylamino-4-methylene coumarin, a photolabile protection group that can be removed upon exposure to 405 nm light. When attached to a lipid’s headgroup, the coumarin moiety temporarily blocks interactions with proteins, preventing the lipid from participating in its normal biological pathways. Upon photoactivation, the coumarin group is removed via a fast radical-based mechanism, restoring the lipid to its native structure and allowing it to engage with its biological targets. Another advantage are the fluorescent properties of the caging group which permits monitoring of cell entry and cell location.

Other photo-activatable groups include nitrobenzyl groups and Bodipy cages. The former often need shorter wavelengths for the (somewhat slower) photo-reaction. In addition, nitrobenzyl groups are not fluorescent and hence cannot not be used for following lipid derivative entry into cells. Bodipy cages have great photo-chemical properties and permit uncaging at longer wavelengths but add significantly more bulk to the lipid derivative.

This uncaging approach is very useful for studying lipid-protein interactions in intact cells and tissue where longer incubation times are needed, as uncaging the lipid in a controlled manner ensures that protein binders recognize the native lipid structure only when and where the researcher intends.

@@

The power of coumarin-caged lipids has been demonstrated in studies examining intracellular lipid signaling. For example, in 2015, Höglinger et al. used a caged sphingosine analog to investigate the role of sphingosine in calcium signaling. They found that upon photoactivation of the caged sphingosine, lysosomal calcium release was triggered, revealing a crucial link between sphingosine metabolism and intracellular calcium dynamics[@hoglingerIntracellularSphingosineReleases2015a]. This discovery significantly expanded the understanding of the sphingosine signaling cascade and its downstream effects.

@@

## Limitations of multi-functionalized lipid probes

*	While significant effort has minimized the effect that each modification has on the biophysical properties of the lipid analogs, it remains the case that adding a diazirine and a terminal alkyne will necessarily affect their physical properties – perhaps affecting sub-cellular localization and binding.

*	The coumarin cage has been repeatedly shown to protect lipid analogs from premature metabolism, but it should be recognized that this comes with some distinct changes to the behavior of the native lipid.
    -	For example, when a cell is exposed to a photocaged lipid probe, the probe intercalates predominantly to endomembranes even if the native lipid is known to reside in the plasma membrane, as was shown for PIP3. This may lead to the identification of non-biologically-relevant hits or require enough time for lipid translocation after uncaging.
*	Similarly, the diazirine and alkyne groups may have minor effects on the biophysical properties of the lipid probe as it interacts with other lipids and proteins within the membrane bilayer. This may be more relevant for proteins which preferentially bind to the acyl chain.
    -	For example, some work has been done to demonstrate that the placement of the diazirine ring along the length of the acyl chain can have impacts on the interactome of the lipid probe[@farleyTrifunctionalFattyAcid2024]. Farley et al (2024) demonstrated this using trifunctional fatty acid probes by placing the diazirine either 3 carbons from the terminus (8-3 FA) or 10 carbons from the terminus (1-10 FA). They identified several proteins with distinct enrichment profiles, including VDAC1 and VDAC2 (which are preferentially enriched to the 8-3 FA probe), and ACAD9 and UBB (which are preferentially enriched to the 1-10 FA probe).
*	Quantification of +UV versus -UV
    -	The details of the quantitative mass spectrometry are described in our Overview of Proteomics Methods. However, we hope to succinctly describe that quantitative mass spectrometry techniques necessarily require normalization at sample input level – thereby exaggerating the abundance of proteins in the -UV control samples.
    -	It should be noted, therefore, that there is little meaning to a protein which is “enriched” toward the -UV condition. For this reason, all proteins which are “depleted” following irradiation with UV have been designated as “no hit” in most of the analyses.
